# Creating New Formats {#new-formats}

R Markdown includes many built-in document and presentation formats which control the style and appearance of the output documents. At their core, these formats are just R functions, and when you include an output format in the YAML front-matter of a document you are really just specifying the format function to call and the parameters to pass to it.

We can create new formats in R Markdown, which makes it easy to customize output formats to use specific options or refer to external files. Defining a new function can be particularly beneficial if you have generated a new template as described in Chapter \@ref(document-templates), as it allows us to use our custom templates without having to copy any files to our local directory.

## Deriving from built-in formats

The easiest way to create a new format is to write a function that calls one of the built-in formats. These built-in formats are designed to be extensible enough to serve as the foundation of custom formats. For example, the following example `quarterly_report` specifies default options for the `html_document`:

```{r, eval=FALSE, tidy=FALSE}
quarterly_report <- function(toc = TRUE) {

  # get the locations of resource files located within the package
  css <- system.file("reports/styles.css", package = "mypackage")
  header <- system.file("reports/quarterly/header.html", package = "mypackage")

  # call the base html_document function
  rmarkdown::html_document(toc = toc,
                           fig_width = 6.5,
                           fig_height = 4,
                           theme = NULL,
                           css = css,
                           includes = includes(before_body = header))
}
```

The new format defined has the following behavior:

1. Provides an option to determine whether there is a table of contents included in the document (implemented by passing toc through to the base format).
2. Sets a default height and width for figures (note this is purposely not user-customizable so as to encourage a standard for all reports of this type).
3. Disables the default bootstrap theme and provides custom CSS in its place.
4. Adds a standard header to every document.

Note that (3) and (4) are implemented using external files that are stored within the package that defines the custom format, so their locations need to be looked up using the  `system.file` function. 

## Fully custom formats

Another lower-level approach is to define a format directly by explicitly specifying knitr options and pandoc command line arguments. At it’s core an R Markdown format consists of:

1. A set of knitr options that govern how Rmd is converted to markdown.
2. A set of pandoc options that govern how markdown is converted to the final output format (e.g. HTML).
3. Some optional flags and filters (typically used to control handling of supporting files).

You can create a new format using the `output_format` function. For example, here is the simplest possible format definition:

```{r, eval=FALSE, tidy=FALSE}
simple_html_format <- function() {
  rmarkdown::output_format(knitr = knitr_options(opts_chunk = list(dev = 'png')),
                           pandoc = pandoc_options(to = "html"),
                           clean_supporting = FALSE)
}
```

Knitr and pandoc options can get considerably more complicated (see the `knitr_options` and `pandoc_options` functions for details). The `clean_supporting` option indicates that you aren't creating self contained output (like a PDF or HTML document with base64 encoded resources) and therefore want to preserve supporting files like plot images generated during knitting.

You can also pass a `base_format` to the `output_format` function if you want to inherit all of the behavior of an existing format but tweak a subset of it’s options.

If there are supporting files required for your format (e.g. HTML/CSS/JS) then you’ll also need to use the other arguments to `output_format` to ensure they are handled correctly (e.g. copying them into place alongside the generated document).

The best way to learn more about creating fully custom formats is to study the source code of the existing built-in formats (e.g. html_document and pdf_document): https://github.com/rstudio/rmarkdown/tree/master/R.  In some cases a custom format will define it’s own pandoc template, which were discussed in Section \@ref(pandoc-templates).

## Using a new format

New formats should be stored within a package and installed onto your local system. This allows the format to be provided to the document YAML. Using our example format `quarterly_report` in the package `mypackage`, we can use this as follows:

```yaml
title: "Habits"
output:
  mypackage::quarterly_report:
    toc: true
```

This says to use the `quarterly_report` function defined in `mypackage` as the output format and to pass `toc = TRUE` as a parameter to the function.
